Modelling the creation of ecological traps by environmentally persistent pathogens

### 2015-05-26 Manuscript/manuscript.tex
Edits to discussion of roles of low/high quality

This work demonstrates that the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.  These impacts are driven largely by the contrasting effects of low and high quality habitat, and their relative balance in the metapopulation.  These contrasting effects derive from differences in colonization and extinction rates, which in turn drive how low and high quality patches interact with and modulate pathogen transmission, either augmenting it in the case of high quality patches, or dampening it in the case of low quality habitat.

When a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.  Attracting colonists helps position high quality patches as infectious "traps", wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).  This increased infection pressure results from both transmission from a persistent environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

The relatively low extinction rate on high quality patches can help to create a stable platform from which the pathogen can spread through the rest of the metapopulation.  Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  Once the pathogen infects high quality patches, the larger number of host colonists produced by these patches allow it to spread to the rest of the metapopulation relatively easily.  This process helps to maintain infectious occupancy throughout the metapopulation, which in turn feeds back on high quality patches to ensure a steady stream of infectious colonists that help maintain the environmental reservoir and the trap effect.  Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread and infectious occupancy in the metapopulation (Fig~\ref{sens}).

In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}). Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation (Fig~\ref{sens}).  This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infectious individuals.  Moreover, the high extinction rate, together with the increased mortality of infectious populations, means that infectious populations do not persist long on low quality patches and the environmental reservoir left behind by these infectious populations may decay before it has the opportunity to infect new susceptible colonists.  As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).

Though there is a consistent trend in the effect of low and high quality habitat on pathogen spread, the net consequences of these effects for total occupancy vary depending on the structure and dynamics of the metapopulation.  Specifically, the findings that low quality habitat inhibits disease spread while high quality habitat enhances it suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.  This work thus expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  Shifting the patch quality distribution towards high quality patches in our model is in many ways analogous to increasing movement in the models of Hess, Gog, and Park, but by including habitat heterogeneity and metapopulation structure, we are able to investigate the role of individual patches and their contribution to total occupancy.

>>>>

This work demonstrates that the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.  
These impacts are driven largely by differences in the colonization and extinction rates of high and low quality habitat, which in turn drive how low and high quality patches interact with and modulate pathogen transmission, either augmenting it in the case of high quality patches, or dampening it in the case of low quality habitat.

Because high quality habitat supports larger host population sizes, host populations on high quality habitat are less susceptible to environmental stochasticity and thus experience fewer extinction events.
Due to this additional stability, high quality habitat is more consistently occupied, by both susceptible and infectious hosts, leading to the increase in infection probability with patch quality.
From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infected hosts, or because susceptible hosts are likely to become infected (either through the environmental reservoir or direct contact).
Thus, when high quality habitat is preferentially selected by migrants (i.e. $\xi_em > 0$), high quality habitats form ecological traps, that is, preferentially selected habitat that reduces fitness relative to other habitats \cite{Robertson2006}).
Moreover, this preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct infection from infectious migrants also selecting high quality habitat, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).

This relatively low extinction rate on high quality patches, and the resulting increased probability of infection, helps then to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
This helps to maintain infectious occupancy throughout the metapopulation, which in turn feeds back on high quality patches to ensure a steady stream of infectious colonists that help maintain the environmental reservoir and the trap effect.  
Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).

In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{endemic}, \ref{popsizes}). 
Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches facilitates susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).  
This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e. low population sizes and high extinction rates) and thus do not support infectious populations for long.
Since they are infrequently colonized (when $\xi_em > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.  
As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{endemic}, \ref{popsizes}).

Though there is a consistent trend in the effect of low and high quality habitat on pathogen spread, the net consequences of these effects for total occupancy vary depending on the traits of the pathogen.  Specifically, the findings that low quality habitat inhibits disease spread while high quality habitat enhances it suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.  This work thus expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  Shifting the patch quality distribution towards high quality patches in our model is in many ways analogous to increasing movement in the models of Hess, Gog, and Park, but by including habitat heterogeneity and metapopulation structure, we are able to investigate the role of individual patches and their contribution to total occupancy.
